Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2021-08-30 write-up/covid-writeup.tex
Updated prevs!
final major paper update

\begin{abstract}
  Coronavirus case-count data has influenced government policies and drives most epidemiological forecasts. Limited testing is cited as the key driver behind minimal information on the COVID-19 pandemic. While expanded testing is laudable, measurement error and selection bias are the two greatest problems limiting our understanding of the COVID-19 pandemic; neither can be fully addressed by increased testing capacity. In this paper, we demonstrate their impact on estimation of point prevalence and the effective reproduction number. We show that estimates based on the millions of molecular tests in the US has the same mean square error as a small simple random sample.  To address this, a procedure is presented that combines case-count data and random samples over time to estimate selection propensities based on key covariate information. We then combine these selection propensities with epidemiological forecast models to construct a \emph{doubly robust} estimation method that accounts for both measurement-error and selection bias.  This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with cumulative demographic information, the United States's only statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.  We end with a series of recommendations based on the proposed methodology.
\end{abstract}

@@

Due to monetary and time constraints, however, random testing is performed infrequently.  As of June 2021, to the best of our knowledge, Indiana is the only state to conduct statewide random sample testing\footnote{This is the only random sample to collect both seroprevalence and diagnostic testing results. The CDC has conducted several seroprevalence studies. See Section~\ref{section:testinginfo} for further discussion.}, which was collected from April 25--29, 2020~\citep{Yiannoutsos2021}.  Such infrequent random testing is likely to provide insufficient information to help researchers and policy makers better understand the disease trajectory which can change rapidly over time.  Therefore, while random testing may be preferable in theory, in practice governments, researchers, and policy makers continue to use coronavirus case-counts to understand the impact of COVID-19 on the population and make data-informed decisions.

@@

 \begin{figure}[!th]
 \centering
 \includegraphics[width=.6\linewidth]{../figs/deaths_ppc.png}
 \caption{Posterior predictive check on aggregate daily death counts.}
 \label{fig:deathsppc}
 \end{figure}

>>>>

\begin{abstract}
  Coronavirus case-count data has influenced government policies and drives most epidemiological forecasts. Limited testing is cited as the key driver behind minimal information on the COVID-19 pandemic. While expanded testing is laudable, measurement error and selection bias are the two greatest problems limiting our understanding of the COVID-19 pandemic; neither can be fully addressed by increased testing capacity. In this paper, we demonstrate their impact on estimation of point prevalence and the effective reproduction number. We show that estimates based on the millions of molecular tests in the US has the same mean square error as a small simple random sample.  To address this, a procedure is presented that combines case-count data and random samples over time to estimate selection propensities based on key covariate information. We then combine these selection propensities with epidemiological forecast models to construct a \emph{doubly robust} estimation method that accounts for both measurement-error and selection bias.  This method is then applied to estimate Indiana's prevalence using case-count, hospitalization, and death data with cumulative demographic information, one of United States's only two statewide random molecular sample collected from April 25--29th, and Facebook's COVID-19 symptom survey.  We end with a series of recommendations based on the proposed methodology.
\end{abstract}

@@

Due to monetary and time constraints, however, random testing is performed infrequently.  As of June 2021, to the best of our knowledge, Indiana and Ohio are the only states to conduct statewide random sample testing\footnote{This is the only random sample to collect both seroprevalence and diagnostic testing results. The CDC has conducted several seroprevalence studies. See Section~\ref{section:testinginfo} for further discussion.}, which was collected from April 25--29, 2020~\citep{Yiannoutsos2021}.  Such infrequent random testing is likely to provide insufficient information to help researchers and policy makers better understand the disease trajectory which can change rapidly over time.  Therefore, while random testing may be preferable in theory, in practice governments, researchers, and policy makers continue to use coronavirus case-counts to understand the impact of COVID-19 on the population and make data-informed decisions.

@@

  \begin{figure}
 \centering
 \begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../figs/deaths_ppc.png}
 \caption{PPC on aggregate daily death counts.}
 \label{fig:deathsppc}
 \end{subfigure}%
 \begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../figs/cases_ppc.png}
 \caption{Posterior on daily case counts.}
 \label{fig:casesppc}
 \end{subfigure}
 \caption{Posterior predictive check (PPC) on death counts and posterior distribution of case counts.}
 \label{fig:ppcs}
 \end{figure}
